---
title: Phytoplankton sampling effort based on OBIS data
date: "`r Sys.Date()`"
author: Pieter Provoost
output: (function(...) {
  rmdformats::robobook(toc_depth = 4, pandoc_args = c("+RTS", "-K2000m", "-RTS"), ...) })
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })  
---

This notebook calculates phytoplankton sampling effort over time for the IOC-UNESCO HAB regions as defined [here](https://github.com/iobis/hab-regions).

## Download data
### Phytoplankton occurrences

First download all phytoplankton occurrences from OBIS using the `robis` package:

```{r message=FALSE, warning=FALSE}
library(robis)
library(dplyr)

if (!file.exists("occurrence.rds")) {
  occ <- occurrence(taxonid = c(17329, 7, 146537), fields = c("decimalLongitude", "decimalLatitude", "date_mid"))
  occ <- occ %>% select(-id)
  saveRDS(occ, file = "occurrence.rds")
} else {
  occ <- readRDS("occurrence.rds")
}
```

### HAB regions shapefile

A shapefile for the HAB regions is available from https://github.com/iobis/hab-regions. 

```{r message=FALSE, warning=FALSE}
library(sf)

if (!file.exists("hab-regions-master")) {
  temp <- tempfile()
  download.file("https://github.com/iobis/hab-regions/archive/refs/heads/master.zip", temp)
  unzip(temp)
  unlink(temp)
}

regions <- read_sf("hab-regions-master/hab_regions/hab_regions.shp", "hab_regions")
```

## Analysis

First calculate years and months, and join the occurrence and HAB region tables:

```{r message=FALSE, warning=FALSE}
library(lubridate)

occ_sf <- occ %>%
  mutate(date = as.POSIXct(date_mid / 1000, origin = "1970-01-01")) %>%
  mutate(year = year(date), month = month(date)) %>%
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326, remove = FALSE) %>%
  st_join(regions, join = st_intersects)
```

Construct a discrete global grid using the [dggridR](https://github.com/r-barnes/dggridR) package (ISEA resolution 7) and assign cells to the occurrences:

```{r message=FALSE, warning=FALSE, results='hide'}
library(dggridR)

dggs <- dgconstruct(projection = "ISEA", res = 8)
occ_sf$cell <- dgtransform(dggs, occ_sf$decimalLatitude, occ_sf$decimalLongitude)

dginfo(dggs)
```

This is a visualization of the grid cells with observations in the European region.

```{r message=FALSE, warning=FALSE, fig.height = 7}
library(rmapshaper)

eur <- regions %>%
  filter(hab_region == 12) %>%
  ms_dissolve() %>%
  ms_simplify()

cells <- occ_sf %>%
  as.data.frame() %>%
  filter(hab_region == 12 & year == 2010) %>%
  group_by(year, month, cell) %>%
  summarize()

grid <- dgcellstogrid(dggs, cells$cell) %>%
  mutate(cell = as.numeric(cell))

cells <- cells %>%
  left_join(grid, on = "cell")

ggplot() +
  geom_sf(data = eur, color = NA, fill = "#bbbbbb") +
  geom_polygon(data = cells, aes(long, lat, group = group), fill = NA, color = "#cc3300") +
  theme(
    panel.background = element_blank(),
    axis.line = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  ) +
  facet_wrap(~paste0(year, "-", sprintf("%02d", month)), ncol = 4)
```

Calculate effort per region and year by counting the number of cell months. The results are written to `effort.csv`.

```{r message=FALSE, warning=FALSE}
library(ggplot2)

region_names <- c("ECA", "CCA", "SAM", "WCA", "ANZ", "SEA", "NAS", "Indian Ocean", "Benguela", "West Africa", "MED", "EUR", "PAC")

stats <- occ_sf %>%
  as.data.frame() %>%
  select(year, month, hab_region, cell) %>%
  filter(!is.na(hab_region)) %>%
  mutate(hab_region = region_names[hab_region]) %>%
  distinct(year, month, hab_region, cell) %>%
  group_by(hab_region, year) %>%
  summarize(effort = n())

write.csv(stats, file = "effort.csv", row.names = FALSE)

ggplot() +
  geom_line(data = stats, aes(x = year, y = effort)) +
  facet_wrap(~hab_region, scales = "free") +
  xlim(c(1985, 2017))
```

